---
title: "Deploying to GitHub Pages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deploying to GitHub Pages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Five things have to be true before a site appears at
`https://you.github.io/yourpkg/`. Four of them are one-time setup; the first
is the one people miss, because the build succeeds without it.

# 1. Declare where the site will live

starlightdown has to know the site's address before it can build a correct
site, because that address determines the path every asset and link is written
against. Set it in `_pkgdown.yml`:

```yaml
url: https://you.github.io/yourpkg
```

or as the first entry of `URL:` in `DESCRIPTION`:

```
URL: https://you.github.io/yourpkg, https://github.com/you/yourpkg
```

The URL is split into an origin and a base path:

| Declared URL | Origin | Base |
| --- | --- | --- |
| `https://you.github.io/yourpkg` | `https://you.github.io` | `/yourpkg` |
| `https://docs.example.org` | `https://docs.example.org` | `/` |
| `https://you.r-universe.dev/yourpkg` | `https://you.r-universe.dev` | `/yourpkg` |

A project page is served from a subdirectory named after the repository, so
the base has to match the repository name. A custom domain serves from the
root, so its base is `/`.

A code-forge URL is deliberately not accepted as a documentation address:
`https://github.com/you/yourpkg` describes where the source lives, not where
the docs are served, and treating it as the site would make the base
`/you/yourpkg`. If nothing usable is declared, the build warns and produces a
site with no canonical URLs or sitemap. It will still work when served from a
root, and break subtly when served from a subdirectory.

# 2. Add the workflow

```{r}
starlightdown::use_starlight_github_actions()
```

This writes `.github/workflows/starlight.yml`, which installs R and your
package's dependencies, installs Quarto, installs the package itself, runs
`build_site()`, then builds the Astro site and uploads it to Pages. Pass
`site_dir` if your site is not in `starlight/`, and `branch` if you publish
from something other than `main`.

The order matters and is not accidental: `build_site()` runs before `npm ci`
because it regenerates `starlight/.starlightdown/`, which contains the
frontend plugin that `package-lock.json` points at.

# 3. Commit the lockfile

`starlight/package-lock.json` must be in the repository. The workflow uses
`npm ci`, which refuses to run without it, and pinning the lockfile is what
makes a CI build reproduce your local one. The scaffold creates it for you;
just do not add it to `.gitignore`.

Commit `starlight/` — its config, `src/content/docs/`, and `public/` — but not
`starlight/node_modules/` or `starlight/dist/`. The scaffold writes a
`.gitignore` that already draws this line. The machine-owned
`starlight/.starlightdown/` directory can be committed or ignored; the workflow
regenerates it either way.

# 4. Enable Pages, with Actions as the source

In the repository, under **Settings → Pages**, set **Source** to
**GitHub Actions**. Not "Deploy from a branch" — the workflow publishes an
artifact directly, and there is no `gh-pages` branch involved.

Or from the command line:

```sh
gh api -X POST repos/OWNER/REPO/pages -f build_type=workflow
```

Pages must be enabled before a deployment can succeed.

# 5. Make the repository public, or pay for it

GitHub Pages serves private repositories only on paid plans. On a free
account the workflow runs to completion and publishes nothing.

# What a deployment looks like

Push to your publishing branch. The workflow builds and deploys, and the site
appears at the URL you declared in step 1. Subsequent pushes redeploy.

Old pkgdown URLs keep working: `/reference/foo.html` is published as a small
redirect page pointing at `/reference/foo/`. These are static files, which is
why the approach works on Pages at all — a `_redirects` file, which some hosts
understand, is inert here.

# When it does not work

**The site loads but every stylesheet and link 404s.** The base path does not
match the repository name. Check `url:`, and remember that renaming the
repository changes the base.

**The deploy job fails with a permissions error.** Pages is not enabled, or is
still set to deploy from a branch.

**`npm ci` fails.** `package-lock.json` is missing, or it disagrees with
`package.json` — most often after switching themes, which adds a dependency.
Run `npm install` in the site directory and commit the updated lockfile. The
vendored plugin is recorded as a path link rather than a version, so upgrading
starlightdown itself does not put the two out of step.

**Vignettes fail to build in CI but work locally.** The workflow installs
Quarto, but a vignette can still depend on a package that is only in your local
library. Anything a vignette needs must be in `Suggests`.

**The build succeeds and the site is stale.** Astro caches rendered content.
The workflow builds from a fresh checkout so this affects local iteration only;
`npm run build -- --force` clears it.

# Other hosts

Nothing here is specific to GitHub Pages except the workflow and the Pages
settings. The build produces a static directory, so any static host serves it.
Set `url:` to wherever it will be served, so that the base path is right, and
publish `starlight/dist/`.
