---
title: "Getting started with starlightdown"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with starlightdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

starlightdown compiles an R package's documentation into an
[Astro Starlight](https://starlight.astro.build/) site. It is a documentation
*compiler*: it reads your package the way pkgdown does, renders it to Markdown,
and hands a typed manifest to a bundled Starlight plugin that does the
presentation.

## What you need

- **R**, obviously, and the package you are documenting **installed**. Examples
  and vignettes are executed against the installed namespace, exactly as
  `example()` and `R CMD check` do.
- **Node.js** (18 or newer), for the Astro build.
- **Quarto**, but only if your package has vignettes. A reference-only site
  builds with no Quarto at all.

## Two commands

```{r}
starlightdown::use_starlight_site() # once
starlightdown::build_site()         # whenever the docs change
```

`use_starlight_site()` creates `starlight/`: an Astro project with the plugin
vendored into `starlight/.starlightdown/plugin/`, pinned dependency versions,
and `npm install` run for you if npm is on the path.

`build_site()` does the compiling. Then:

```sh
cd starlight
npm run dev    # preview at localhost:4321
npm run build  # production build into starlight/dist
```

## What gets built

| Source | Becomes |
| --- | --- |
| `man/*.Rd` | `/reference/<topic>/`, with executed examples and figures |
| `_pkgdown.yml` `reference:` | the grouped function index at `/reference/` |
| `vignettes/*.Rmd`, `*.qmd` | `/articles/<name>/`, executed by Quarto |
| `README.md` | the home page |
| `NEWS.md` | `/news/` |
| `inst/CITATION` | the citation block on the home page |

Old pkgdown URLs keep working: every `/reference/foo.html` gets a redirect page
pointing at `/reference/foo/`.

## `_pkgdown.yml` stays canonical

If you already have a pkgdown site, there is nothing to convert. starlightdown
reads the same file:

- `url:` sets the site origin and the base path.
- `reference:` sections, including `starts_with()` and friends, group the
  function index and order the sidebar.
- `articles:` sections order the article sidebar.
- `redirects:` entries become redirect pages.

`migrate_from_pkgdown()` does not rewrite anything. It scaffolds the site if
needed and prints what carries over and what has no equivalent — custom navbar
components, pkgdown themes, HTML includes:

```{r}
starlightdown::migrate_from_pkgdown()
```

You can keep both generators running side by side.

## Options worth knowing

```{r}
starlightdown::build_site(
  articles = FALSE,     # skip vignettes (and the Quarto requirement)
  examples = FALSE,     # show example code without running it
  run_dont_run = TRUE,  # execute \dontrun{} blocks too
  lazy = FALSE,         # ignore Quarto's freeze cache and re-execute
  theme = "nova"        # "default" (bundled) or "nova"
)
```

starlightdown-specific settings live under a `starlightdown:` key in
`_pkgdown.yml`, so there is still only one configuration file:

```yaml
url: https://you.github.io/yourpkg

starlightdown:
  theme: default
  quickstart: |
    library(yourpkg)
    do_the_thing()
```

## How a build behaves

The whole content tree is written to a staging directory, validated, and only
then swapped into place. Two things follow from that, and both are deliberate:

- A build that fails validation leaves the published site untouched. You never
  ship a half-written tree.
- A page whose source you deleted cannot survive as a stale file, because the
  directory is replaced rather than written over.

The validation gates are the ones that catch real breakage: every page's
frontmatter parses and has a title, every relative image exists next to the
page that names it, every site-absolute link matches a route or a redirect, and
routes and files are the same set.

Builds are deterministic. Building twice without changing anything produces
byte-identical output, which makes the generated site reviewable in a diff.

## Editing the result

Everything under `starlight/.starlightdown/` is machine-owned and rewritten on
every build. Everything else is yours:

- `starlight/astro.config.mjs` — add Astro integrations here.
- `starlight/src/styles/custom.css` — loaded after the theme, so your rules win.
- `starlight/package.json` — starlightdown edits dependencies through
  `jsonlite`, never by rewriting text.
